---
title: "Single-cell RNA-seq data analysis, HTO demultiplexing"
author: "Dresden-concept Genome Center, TU Dresden"
date: "May 2020"
geometry: margin=2cm
output:
   html_document:
    toc: true
    highlight: tango
    theme: paper
    code_folding: hide
---

```{r setup, message=FALSE, warning=FALSE}
# R Options
options(stringsAsFactors=FALSE)

# Required libraries
library(Seurat) # main
library(ggplot2) # plots
library(patchwork) # combination of plots
library(magrittr) # %>% operator

# Other libraries we use
# Knit: knitr
# Data handling: dplyr, tidyr, purrr
# Tables: kableExtra
# Plots: ggsci

# Source plotting functions
source("R/functions_io.R")
source("R/functions_plotting.R")
source("R/functions_analysis.R")
source("R/functions_util.R")

# Knitr default options
knitr::opts_chunk$set(echo=TRUE, cache=FALSE, message=FALSE, warning=FALSE, fig.width=10)
```

# Dataset description
* Peripheral blood mononuclear cells (PBMCs) from 8 different donors
* Cells from each donor are labeled with a hashing antibody
* Samples were subsequently pooled and run on a single lane of the 10X Chromium v2 system
* Dataset and description taken from the Seurat HTO vignette

## Project-specific parameters
Open this code chunk to read all parameters that are set specifically for your project. 
```{r project_parameters, class.source='fold-show'}
param = list()

# Project ID
param$project = "HTO_testDataset"

# Input data path in case Cell Ranger was run 
param$path_data = "test_datasets/10x_pbmc_hto_GSE108313/counts"

# Output path
param$path_out = "test_datasets/10x_pbmc_hto_GSE108313/demultiplexed"
if (!file.exists(param$path_out)) dir.create(param$path_out, recursive=TRUE)

# HTO names
#   HTOs have an ID that is included in the 'features.tsv' input file
#   We additionally ask for readable names that are used throughout this report 
#   This could look like this, where HTO1-3 are the IDs included raw dataset
#     param$hto.names = setNames(c("NameA", "NameB", "NameC"), c("HTO1", "HTO2", "HTO3"))
param$hto_names = setNames(c("htoA","htoB","htoC","htoD","htoE","htoF","htoG","htoH"), 
                           c("htoA","htoB","htoC","htoD","htoE","htoF","htoG","htoH")) # Name equals ID in this test case

# Prefix of mitochondrial genes 
param$mt = "^MT-"

# Main color to use for plots
param$col = "palevioletred"

# Sample data to at most n cells (mainly for tests); set to NULL to deactivate
param$sample_cells = NULL
```

```{r hto_colours}
param$col_hto_global = ggsci::pal_npg()(3)
param$col_hto_all = ggsci::pal_npg()(factorial(length(param$hto_names)) + 1)
param$col_hto_collapsed = ggsci::pal_npg()(length(param$hto_names) + 2)
hto_levels = c("Negative", "Doublet", param$hto_names[length(param$hto_names):1])
names(param$col_hto_collapsed) = hto_levels
```

# Read input data
In this first section of the report, we read 10X data from the files produced by CellRanger:  

* **barcodes.tsv.gz**: All cell barcodes  
* **features.tsv.gz**: (Ensembl) ID, name, and type for each feature  
* **matrix.mtx.gz**: Counts for all features  

and setup a Seurat object. This object includes different data types in separate assays:  

* **Gene Expresssion** in assay **RNA**  
* **Antibody Capture** in **HTO** and **ADT**,   
* **CRISPR Guide Capture** in **Crispr**, and  
* **Custom** in **Custom**.   

Note that 'Antibody Capture' features can correspond to 'HTO' and 'ADT' and can be distinguised based on provided HTO names. 
```{r hto_read}
# Load the dataset with its assays and create a Seurat object; pass hto_names so that HTO will be a separate assay
sc = Read10xDataset(param$path_data, hto_names=param$hto_names)

# If requested: sample at most n cells
if (!is.null(param$sample_cells)) {
  sampled_barcodes = sample(Seurat::Cells(sc), min(param$sample_cells, length(Seurat::Cells(sc))))
  sc = subset(sc, cells=sampled_barcodes)
}

# Discard cells in other assays which have no HTO counts 
original_assay_names = setdiff(Seurat::Assays(sc), "HTO") 
assay_cells_hto_zero = Seurat::GetAssayData(sc, assay="HTO", slot="counts") %>%
  as.data.frame %>%
  colSums() == 0 
if (sum(assay_cells_hto_zero)>0) {
  sc = subset(sc, cells=which(assay_cells_hto_zero))
  warning("Discarded ", sum(assay_cells_hto_zero), " cells with 0 HTO counts.")
}

# Remember original assay names
original_assay_names = setdiff(Seurat::Assays(sc), "HTO") 
```

```{r hto_cells_summary}
summary = purrr::map_dfr(original_assay_names, function(a) {
  data.frame(Assay=a,
             CellsTotal=length(assay_cells_hto_zero),
             CellsWithHtoReads=sum(!assay_cells_hto_zero),
             CellsWithoutHtoReads=sum(assay_cells_hto_zero))
  })

knitr::kable(summary, align="l", caption="Dataset summary") %>% 
  kableExtra::kable_styling(bootstrap_options=c("striped", "hover"), full_width=FALSE, position="left")
```

# Demutliplexing with hashtag oligos (HTOs)
This section of the report shows how cells are assigned to their sample-of-origin. 

## Normalisation of HTO counts
We start the analysis by normalising raw HTO counts. HTO counts for each cell are divided by the total counts for that cell and multiplied by 10,000. This is then natural-log transformed. 
```{r hto_norm}
sc = Seurat::NormalizeData(sc, assay="HTO", normalization.method="LogNormalize")

# alternative: CLR
# sc = NormalizeData(sc, assay="HTO", normalization.method="CLR") 
```

## Classification of cells based on normalised HTO data
We assign cells to their sample-of-origin, annotate negative cells that cannot be assigned to any sample, and doublet cells that are assigned to two samples. 
```{r hto_demux, results="asis"}
# Demultiplex HTOs
sc = Seurat::HTODemux(sc, assay="HTO", positive.quantile=0.99)
  
# Sort Idents levels for nicer plotting
Seurat::Idents(sc) = factor(Seurat::Idents(sc), levels=hto_levels)
sc$hash.ID = factor(sc$hash.ID, levels=hto_levels)
  
# HTO classification results
hash_ID_table = sc@meta.data %>% 
  dplyr::count(hash.ID) %>% 
  dplyr::rename(HTO=hash.ID) %>% 
  dplyr::mutate(Perc=round(n/sum(n)*100,2)) %>% 
  as.data.frame

p1 = ggplot(sc@meta.data %>% dplyr::count(HTO_classification.global), 
            aes(x="", y=n, fill=HTO_classification.global)) + 
  geom_bar(width=1, stat="identity") + 
  coord_polar("y", start=0) + 
  xlab("") + ggplot2::ylab("")
p1 = PlotMystyle(p=p1, title="HTO global classification results", col=param$col_hto_global)
  
p2 = ggplot(sc@meta.data %>% dplyr::count(hash.ID), 
            aes(x="", y=n, fill=hash.ID)) + 
  geom_bar(width=1, stat="identity") + 
  coord_polar("y", start=0) + 
  xlab("") + ylab("")
p2 = PlotMystyle(p=p2, title="HTO classification results", col=param$col_hto_collapsed)

p = p1 + p2 + 
  patchwork::plot_annotation(title="HTO classification results") + 
  gridExtra::tableGrob(hash_ID_table, rows=NULL)
p
```

# Visualisation of raw and normalised HTO data {.tabset}
This section of the report visualises raw and normalised HTO data to understand whether the demultiplexing step has worked well. 
```{r hto_plot_counts_raw}
# Distribution of HTO counts before and after normalisation
hto_t_raw = Seurat::GetAssayData(sc, assay="HTO", slot="counts") %>% 
  as.data.frame %>% t %>% as.data.frame
hto_t_raw_pseudo = hto_t_raw + 1
hto_t_norm = Seurat::GetAssayData(sc, assay="HTO", slot="data") %>% 
  as.data.frame %>% t %>% as.data.frame

p1 = ggplot(hto_t_raw_pseudo %>% tidyr::pivot_longer(everything()), aes(x=name, y=value, fill=name)) + 
  geom_violin() + 
  scale_y_continuous(trans="log2") + 
  xlab("") + ylab("") 
p1 = PlotMystyle(p1, title="HTO raw counts", col=param$col_hto_collapsed, legend_title="HTO")
  
p2 = ggplot(hto_t_norm %>% tidyr::pivot_longer(tidyr::everything()), 
            aes(x=name, y=value, fill=name)) + 
  geom_violin() + 
  xlab("") + ylab("")
p2 = PlotMystyle(p2, title="HTO normalised counts", col=param$col_hto_collapsed, legend_title="HTO")

p = p1 + p2 & theme(legend.position="bottom") 
p = p + patchwork::plot_annotation("HTO counts before and after normalisation") + 
  patchwork::plot_layout(guides = "collect")
p
```

Pairs of raw (left tab) and normalised (right tab) HTO counts are visualised to confirm mutal exclusivity in singlet cells. Data points correspond to measured HTO counts per HTO, colours correspond to the assigned samples-of-origin. 

## Raw HTO data
```{r hto_plot_counts_norm, fig.height=10}
n = DfAllColumnCombinations(x=hto_t_raw_pseudo, cell_classification=sc$hash.ID)

# plot
p = ggplot(n, aes(x=value1, y=value2, color=cell_classification)) + 
  geom_point() + 
  scale_x_continuous(trans="log2") + scale_y_continuous(trans="log2") + 
  scale_color_manual(values=param$col_hto_collapsed)
p = PlotMystyle(p)
p = p + facet_grid(name2~name1, drop=FALSE) + 
  theme(axis.title.x=element_blank(), strip.text.x=element_text(size=10, color="black"),
        axis.title.y=element_blank(), strip.text.y=element_text(size=10, color="black"),
        strip.background = element_rect(colour="white", fill="lightgrey"),
        legend.position="bottom", 
        axis.text.x=element_text(angle=45, hjust=1, vjust=0.5)) + 
  patchwork::plot_annotation("Raw HTO counts")
p
```

## Normalized HTO data
```{r hto_plot_norm_scatter, fig.height=10}  
n = DfAllColumnCombinations(x=hto_t_norm, cell_classification=sc$hash.ID)
p = ggplot(n, aes(x=value1, y=value2, color=cell_classification)) + 
  geom_point() + 
  scale_x_continuous(trans="log2") + scale_y_continuous(trans="log2") + 
  scale_color_manual(values=param$col_hto_collapsed)
p = PlotMystyle(p)
p = p + facet_grid(name2~name1, drop=FALSE) + 
  theme(axis.title.x=element_blank(), strip.text.x=element_text(size=10, color="black"),
        axis.title.y=element_blank(), strip.text.y=element_text(size=10, color="black"),
        strip.background = element_rect(colour="white", fill="lightgrey"),
        legend.position="bottom") + 
  patchwork::plot_annotation("Normalised HTO data")
p
```

The following ridge plots visualise the enrichment of assigned sample-of-origin for the respective normalised HTO counts. 
```{r hto_plot_norm, fig.height=10}
# Group cells based on HTO classification 
p = RidgePlot(sc, assay="HTO", features=rownames(Seurat::GetAssay(sc, assay="HTO")), 
              same.y.lims=TRUE, cols=param$col_hto_collapsed, combine=FALSE)
for (i in 1:length(p)) p[[i]] = PlotMystyle(p[[i]], legend_title="Classified cells")
p = patchwork::wrap_plots(p, ncol = 2) + 
  patchwork::plot_annotation("Normalised HTO data") + 
  patchwork::plot_layout(guides = "collect") & 
  theme(legend.position="bottom") 
p
```

Lastly, we compare the number of features between classified cells. 
```{r hto_plot_features}
# Number of features in the different cells
nfeature_metrics = grep("_HTO", 
                        grep("nFeature_", colnames(sc[[]]), v=TRUE),
                        v=TRUE,
                        invert=TRUE)
p = VlnPlot(sc, features=nfeature_metrics, idents=levels(Seurat::Idents(sc)), ncol=3, pt.size=0) + 
  geom_violin(color=NA) +
  xlab("") 

p = PlotMystyle(p, 
                title="Number of features for HTO-classified cells", 
                col=param$col_hto_collapsed, 
                legend_title="Classified cells")
p
```

# Remove cells classified as doublet or negative
This section of the report states the number of cells that remain after negative and doublet cells are removed. 
```{r hto_subset}
sc.all = sc
sc = subset(sc, idents=c("Negative", "Doublet"), invert=TRUE)
sc
```

# Preliminary visualisation of demultiplexed RNA data
This section of the report provides first insights into your RNA dataset based on a preliminary pre-processing of the RNA data using the standard scRNA-seq workflow. 
```{r prelim_analysis, results='hide'}
sc.all = prelim.analysis(sc=sc.all, mt=param$mt, pc_n=10)
sc = prelim.analysis(sc=sc, mt=param$mt, pc_n=10)
```

## Visualisation with UMAP {.tabset}
We use a UMAP to visualise and explore a dataset. The goal is to place similar cells together in 2D space, and learn about the biology underlying the data. Cells are color-coded according to the assigned sample-of-origin. 

Take care not to mis-read a UMAP:  

* Parameters influence the plot (we use defaults here)  
* Cluster sizes relative to each other mean nothing, since the method has a local notion of distance  
* Distances between clusters might not mean anything  
* You may need more than one plot  
  
For a nice read to intuitively understand UMAP, see https://pair-code.github.io/understanding-umap/. 

### Before filtering
```{r umap_before}
p = Seurat::DimPlot(sc.all, reduction="umap", group.by="hash.ID", cols=param$col.hto.all) 
p = PlotMystyle(p, 
                title="UMAP, cells coloured by HTO classification, including doublets and negatives", 
                legend_title="Cell classification", 
                legend_position="bottom")
p
```

### After filtering
```{r umap_after}
# Plot UMAP after HTO filtering
p = Seurat::DimPlot(sc, reduction="umap", group.by="HTO_classification", cols=param$col_hto_collapsed) 
p = PlotMystyle(p, 
                title="UMAP, cells coloured by HTO classification, singlets only", 
                legend_title="Cell classification", 
                legend_position="bottom")
p
```

# Write out demultiplexed data
Finally, demultiplexed RNA data are written back to file. 
```{r hto_save_samples}
# Create a directory for the demultiplexed data
demux_base_dir = file.path(param$path_out, "demultiplexed_samples")
dir.create(demux_base_dir, showWarnings=FALSE)
demux_samples_paths = c()

# Save each sample in a separate directory
samples = levels(Seurat::Idents(sc))
for (s in samples) {
  p = ExportSeuratAssayDataToDir(sc[,Seurat::Idents(sc)==s], 
                                 dir=file.path(demux_base_dir, s), 
                                 assays=original_assay_names, 
                                 slot="counts",
                                 include_cell_metadata_cols=c("HTO_maxID", 
                                                              "HTO_classification", 
                                                              "HTO_classification.global", 
                                                              "hash.ID"))
  demux_samples_paths = c(demux_samples_paths, p)
}

message(demux_samples_paths)
```

