---
author:
  - "[Dresden-concept Genome Center, TU Dresden](https://genomecenter.tu-dresden.de/)"
  - "[Research Core Unit Genomics, Hannover Medical School](https://www.mhh.de/genomics)"
date: "`r format(Sys.time(), '%B, %Y')`"
geometry: margin=2cm
output:
   html_document:
    toc: true
    toc_depth: 4
    code_folding: hide
    highlight: tango
    theme: paper
params:
   # Project ID
   project_id: 10x_pbmc_5k 
   
   # Input directory
   path_data: test_datasets/10x_pbmc_5k_protein/counts/
  
   # Metrics summary
   stats: test_datasets/10x_pbmc_5k_protein/metrics_summary.csv
   
   # Output directory
   path_out: test_datasets/10x_pbmc_5k_protein/counts/
   
   # Normalisation: LogNormalize or SCTransform
   norm: "LogNormalize"
   
   # Prefix of mitochondrial genes
   mt_names: "^MT-"
   
   # Filter for low-quality cells
   cell_filter: !r list(nFeature_RNA=c(200, NA), percent_mt=c(NA, 20), nCount_RNA=c(1000, NA))
   
   # The number of PCs to use
   pc_n: 10
   
   # Main color to use for plots
   col: "palevioletred"
   
   # Downsample data to at most N cells (mainly for tests); set to NULL to deactivate
   downsample_cells_n: NULL
   
   # Path to git
   path_to_git: "."
   
   # Debugging mode: 
   # 'default_debugging' for default, 'terminal_debugger' for debugging without X11, 'print_traceback' for non-interactive sessions 
   debugging_mode: on_error_default_debugging
title: "Single-cell RNA-seq data analysis of project `r params$project_id`"
subtitle: "Identification of multiplets"
---

```{r setup, message=FALSE, warning=FALSE}
# R Options
options(stringsAsFactors=FALSE,
        "citation_format"="pandoc", 
        dplyr.summarise.inform=FALSE, 
        knitr.table.format="html",
        future.globals.maxSize=2000000000, mc.cores=4, 
        future.fork.enable=TRUE, future.plan="multicore",
        future.rng.onMisuse="ignore")

# Required libraries
library(Seurat) # main
library(ggplot2) # plots
library(patchwork) # combination of plots
library(magrittr) # %>% operator

# Other libraries we use
# Knit: knitr
# Data handling: dplyr, tidyr, purrr
# Tables: kableExtra
# Plots: ggsci

# Knitr default options
knitr::opts_chunk$set(echo=TRUE,                     # output code
                      cache=FALSE,                   # do not cache results
                      message=TRUE,                  # show messages
                      warning=TRUE,                  # show warnings
                      tidy=FALSE,                    # do not auto-tidy-up code
                      fig.width=10,                  # default fig width in inches
                      class.source='fold-hide',      # by default collapse code blocks
                      dev=c('png', 'pdf'),           # create figures in png and pdf; the first device (png) will be used for HTML output
                      dev.args=list(png=list(type="cairo"),  # png: use cairo - works on cluster, supports anti-aliasing (more smooth)
                                    pdf=list(bg="white")),     # pdf: use cairo - works on cluster, supports anti-aliasing (more smooth)
                      dpi=96,                        # figure resolution (note: final = dpi*fig.retina; will not work with pandoc see below)
                      fig.retina=2                   # retina multiplier for HTML output (note: final = dpi*fig.retina)
                                     
)
```

```{r preparation_and_initial_checks}
# Path for figures in png and pdf format
knitr::opts_chunk$set(fig.path=paste(params$path_out, "figures/", sep="/"))


# Git directory and files to source must be done first, then all helper functions can be sourced
git_files_to_source = c("R/functions_io.R",
              "R/functions_plotting.R",
              "R/functions_util.R")
git_files_to_source = paste(params$path_to_git, git_files_to_source, sep="/")
file_exists = purrr::map_lgl(git_files_to_source, file.exists)

if (any(!file_exists)) stop(paste("The following files could not be found:",paste(git_files_to_source[!file_exists], collapse=", "), ". Please check the git directory at '", params$path_to_git, "'.!"))
invisible(purrr::map(git_files_to_source, source))

# Debugging mode: 
# 'default_debugging' for default, 'terminal_debugger' for debugging without X11, 'print_traceback' for non-interactive sessions 
invisible(switch (params$debugging_mode, 
        default_debugging=on_error_default_debugging(), 
        terminal_debugger=on_error_start_terminal_debugger(),
        print_traceback=on_error_just_print_traceback(),
        on_error_default_debugging()))

# Set output hooks
knitr::knit_hooks$set(message=format_message, warning=format_warning)

# Create output directories
if (!file.exists(params$path_out)) dir.create(params$path_out, recursive=TRUE, showWarnings=FALSE)

# Do checks
error_messages = c()
# Check installed packages
error_messages = c(error_messages, check_installed_packages())
# Check parameters TODO
#error_messages = c(error_messages, check_parameters(param))

# Stop here if there are errors
if (length(error_messages)) stop(paste(c("", paste("*", error_messages)), collapse="\n"))
```

# Read input data
In this first section of the report, we begin by printing mapping statistics that have been produced prior to this workflow.
```{r mapping_stats, message=TRUE, results='asis'}
# Are statistics provided?
if (!is.null(params$stats)) {
  cat('\n## Stats {.tabset}\n\n')
  mapping_stats = read.delim(params$stats,sep=",", header=FALSE, check.names=FALSE) %>% t() %>% as.data.frame()
  colnames(mapping_stats) = c("V1","Dataset")
  rownames(mapping_stats) = mapping_stats$V1
  mapping_stats$V1 = NULL

  mapping_stats_general = mapping_stats[!grepl("^Antibody:",rownames(mapping_stats)),,drop=FALSE]
  mapping_stats_antibodies = mapping_stats[grepl("^Antibody:",rownames(mapping_stats)),,drop=FALSE]
  
  cat('\n### General\n\n')
  print(knitr::kable(mapping_stats_general, align="l", caption="General statistics") %>% 
  kableExtra::kable_styling(bootstrap_options=c("striped", "hover")))
 
  cat('\n### Antibodies\n\n')
  print(knitr::kable(mapping_stats_antibodies, align="l", caption="Antibody statistics") %>% 
  kableExtra::kable_styling(bootstrap_options=c("striped", "hover")))
} else { 
  message("Mapping statistics cannot be shown. No valid file provided.")
}
```

# Filtering and normalisation/scaling
Then, we read 10X data from the files produced by CellRanger:  

* **barcodes.tsv.gz**: All cell barcodes  
* **features.tsv.gz**: (Ensembl) ID, name, and type for each feature  
* **matrix.mtx.gz**: Counts for all features  

and setup a Seurat object with the **Gene Expresssion** data in assay **RNA**.

```{r read_datasets}
# Load the dataset with its assays and create a Seurat object with the **Gene Expresssion** data in assay **RNA**.
sc = ReadSparseMatrix(params$path_data, project=params$project_id, row_name_column=1)
sc = sc[[1]]

# If requested: sample at most n cells
if (!is.null(params$downsample_cells_n)) {
  sampled_barcodes = sample(Seurat::Cells(sc), min(params$downsample_cells_n, length(Seurat::Cells(sc))))
  sc = subset(sc, cells=sampled_barcodes)
}

# Calculate percentage of counts in mitochondrial genes
mt_features = grep(pattern=params$mt_names, rownames(sc), value=TRUE)
sc = Seurat::PercentageFeatureSet(sc, features=mt_features, col.name="percent_mt", assay="RNA")

# Calculate percentage of counts in ribosomal genes
ribo_features = grep(pattern="^RP[SL]", rownames(sc), value=TRUE, ignore.case=TRUE)
sc = Seurat::PercentageFeatureSet(sc, features=ribo_features, col.name="percent_ribo", assay="RNA")

# Calculate percentage of counts in ERCC for each Seurat object (if assay is available)
if ("ERCC" %in% Seurat::Assays(sc)) {
  sc$percent_ercc = sc$nCount_ERCC/(sc$nCount_ERCC + sc$nCount_RNA)*100
}
```

To ensure that input data is cleared of low-quality cell clusters, we first threshold cells according to RNA nCount, nFeature and percent_mt. Cells filtered based on the following thresholds: 
```{r filter_print_cutoffs}
cell_filter_lst = params$cell_filter
is_numeric_filter = purrr::map_lgl(cell_filter_lst, function(f) return(is.numeric(f) & length(f)==2))

# Numeric cell filters
if (length(cell_filter_lst[is_numeric_filter]) > 0) {
  purrr::invoke(rbind, cell_filter_lst[is_numeric_filter]) %>%
    knitr::kable(align="l", caption="Numeric filters applied to cells", col.names=c("Min", "Max")) %>% 
    kableExtra::kable_styling(bootstrap_options=c("striped", "hover"))
}

# Categorial cell filters
if (length(cell_filter_lst[!is_numeric_filter]) > 0) {
purrr::invoke(rbind, cell_filter_lst[!is_numeric_filter] %>% purrr::map(paste, collapse=",")) %>%
  knitr::kable(align="l", caption="Categorial filters applied to cells", col.names=c("Values")) %>% 
  kableExtra::kable_styling(bootstrap_options=c("striped", "hover"))
}
```

The number of excluded cells and features is as follows: 
```{r filter_cells}
# Iterate over datasets and filters
# Record a cell if it does not pass the filter
sc_cells_to_exclude = purrr::map(list_names(params$cell_filter), function(f) {
  filter = params$cell_filter[[f]]
  if (is.numeric(filter)) {
    if (is.na(filter[1])) filter[1] = -Inf # Minimum
    if (is.na(filter[2])) filter[2] = Inf  # Maximum 
    idx_exclude = sc[[f, drop=TRUE]] < filter[1] | sc[[f, drop=TRUE]] > filter[2]
    return(names(which(idx_exclude)))
  } else if (is.character(filter)) { 
    return(names(which(sc[[f, drop=TRUE]] %in% filter)))
  }
})


# Summarise
sc_cells_to_exclude_summary = purrr::map_dfr(sc_cells_to_exclude, function(s) {
  return(data.frame(cells=length(s)))
})

rownames(sc_cells_to_exclude_summary) = names(sc_cells_to_exclude)
knitr::kable(sc_cells_to_exclude_summary, align="l", caption="Number of excluded cells") %>% 
  kableExtra::kable_styling(bootstrap_options=c("striped", "hover")) 

# Now filter
cells_to_keep = Cells(sc)
cells_to_keep = cells_to_keep[!cells_to_keep %in% purrr::flatten_chr(sc_cells_to_exclude)]
if (length(cells_to_keep)==0) stop("Error: no cells remaining after filtering!")
sc = subset(sc, cells=cells_to_keep)
```

After filtering, the size of the Seurat object is: 
```{r filter_size_after}
sc
```

Then, we normalise, identify genes that are highly variable between cells and scale those genes. There are two types of normalisation strategy available:

1. A  __standard log normalisation__, where counts for each cell are divided by the total counts for that cell and multiplied by 10,000. This is then natural-log transformed.   
2. __SCTransform__, a new and more sophisticated normalisation method which includes normalisation, variable genes and scaling. 

```{r normalisation_and_scaling, results="hide", warning=FALSE}
if (params$norm == "LogNormalize") {
  # Standard normalisation
  sc = Seurat::NormalizeData(sc, normalization.method="LogNormalize", scale.factor=10000, verbose=FALSE)
  
  # Find variable features from normalised data
  sc = Seurat::FindVariableFeatures(sc, selection.method="vst", nfeatures=3000, verbose=FALSE)
  
  # Scale (only variable features) 
  sc = Seurat::ScaleData(sc, Seurat::VariableFeatures(sc), verbose=FALSE, vars.to.regress=c())
} else if (params$norm == "SCTransform") {
  
  # Run SCTransform
  sc = SCTransform(sc, vars.to.regress=c(), min_cells=3, verbose=FALSE, return.only.var.genes=TRUE, method=ifelse(packages_installed("glmGamPoi"), "glmGamPoi", "poisson"))
}
```

# Dimensionality reduction, clustering and visualisation



```{r dimred_and_visualisation, results="hide", warning=FALSE}
# Run PCA
sc = Seurat::RunPCA(sc, features=Seurat::VariableFeatures(object=sc), verbose=FALSE)

# Choose the number of PCs to use for clustering
sc = Seurat::FindNeighbors(sc, dims=1:params$pc_n, verbose=FALSE)

# Find clusters using Seurats standard algorithm
sc = Seurat::FindClusters(sc, resolution=0.5, algorithm=1)

# Run UMAP
sc = Seurat::RunUMAP(sc, dims=1:params$pc_n, verbose=FALSE, umap.method="uwot")
```

Next, we use Principal Component Analysis (PCA)

A single-cell dataset of 20,000 genes and 5,000 cells has 20,000 dimensions. At this point of the analysis, we have already reduced the dimensionality of the dataset to 3,000 variable genes. The biological manifold however can be described by far fewer dimensions than the number of (variable) genes. Dimension reduction methods aim to find these dimensions. There are two general purposes for dimension reduction methods: to summarise a dataset, and to visualise a dataset. 

We use Principal Component Analysis (PCA) to summarise a dataset, overcoming noise and reducing the data to its essential components. Each principal component (PC) represents a "metafeature" that combines information across a correlated gene set. Later, we use Uniform Manifold Approximation and Projection (UMAP) to visualise the dataset, placing similar cells together in 2D space, see below. 

To decide how many PCs to include in downstream analyses, we visualise the cells and genes that define the PCA.
```{r pca}
# Run PCA for default normalisation
sc = Seurat::RunPCA(sc, features=Seurat::VariableFeatures(object=sc), verbose=FALSE, npcs=min(50, ncol(sc)))

p_list = Seurat::VizDimLoadings(sc, dims=1:2, reduction="pca", col=param$col, combine=FALSE)
for (i in seq(p_list)) p_list[[i]] = p_list[[i]] + AddStyle()
p =  patchwork::wrap_plots(p_list, ncol=2) + patchwork::plot_annotation("Top gene loadings of the first two PCs") 
p

p = Seurat::DimPlot(sc, reduction="pca", cols=param$col_samples) + 
  AddStyle(title="Cells arranged by the first two PCs", legend_position="bottom")
p

p = Seurat::DimHeatmap(sc, dims=1:2, cells=min(500, ncol(sc)), balanced=TRUE, fast=FALSE)
p
```


# Doublet detection

```{r doublet_detection}

# Identification of optimal pK
sweep.res.list_sc = DoubletFinder::paramSweep_v3(sc, PCs=1:params$pc_n, sct=params$norm == "SCTransform")
sweep.stats_sc = DoubletFinder::summarizeSweep(sweep.res.list_sc, GT=FALSE)
bcmvn_sc = DoubletFinder::find.pK(sweep.stats_sc)

# Estimate the number of loaded cells
num_loaded_cells = 2*ncol(sc)

# Estimate the doublet rate: 10x says 0.8% / 1000 loaded cells
exp_doublet_rate = num_loaded_cells/1000*0.008

# Estimated number of doublets based on poison distribution
nExp_poi = round(0.15*length(Cells(sc)))

# Run DoubletFinder
sc = DoubletFinder::doubletFinder_v3(sc, PCs=1:params$pc_n, pN=0.25, pK=0.25, nExp=exp_doublet_rate, reuse.pANN=FALSE, sct=params$norm == "SCTransform")
```

# Output results

```{r}

```



# Parameter table
The following parameters were used to run the workflow.  
```{r parameters_table}
out = ScrnaseqParamsInfo(params=params)

knitr::kable(out, align="l") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), full_width=FALSE, position="left")
```

# Software versions
This report was generated using the [scrnaseq](https://github.com/ktrns/scrnaseq) GitHub repository. Software versions were collected at run time. 
```{r versions}
out = ScrnaseqSessionInfo(params$path_to_git)

knitr::kable(out, align="l") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), full_width=FALSE, position="left")
```
